---
title: "Homework 1"
author: "Spencer Matthews"
date: "10/11/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(dplyr)
```


## Problem 1

Complete the following programming exercises in any language of your choice.

**(a) Discrete time individual-based modeling**

*(i) Write a function that simulates a fixed number of generations of a discrete time individual-based SIR model. Your function should have the following user specified arguments: initial vector of individual labels, infection probability $p$, and number of generations $r$*

```{r}
#' Individual Discrete-Time SIR Model
#'
#' @param init_labels The initial labels of the population, a character vector of S,I,R
#' @param p The infection probability, a number between 0 and 1
#' @param r The number of generations (ie, the number of time steps), an integer
individual_discrete <- function(init_labels, p, r) {
  step_labs <- init_labels
  res <- data.frame(step_0 = init_labels)
  for (i in 1:r) {
    i_t <- sum(step_labs == "I")
    step_prob <- 1 - (1-p)^i_t
    step_labs <- purrr::map_chr(
      .x = step_labs,
      .f = ~{
        if (.x == "S") {
          sample(c("I", "S"), size = 1, prob = c(step_prob, 1 - step_prob))
        } else{
          "R"
        }
      }
    )
    res <- cbind(res, data.frame(step_labs))
  }
  res %>%
    magrittr::set_colnames(
      stringr::str_c("step_", 0:r)
    ) %>%
    magrittr::set_rownames(
      stringr::str_c("person_", 1:length(init_labels))
    )
}
```


*(ii) Assuming that initially the population of interest has 9 susceptible individuals, 1 infectious individual, and 0 removed individuals, print 5 realizations of label vectors at each generation using $p = 0.3$ and $r = 5$.*

```{r}
init <- c("S", "S", "S", "S", "S", "S", "I", "S", "S", "S")

for (i in 1:5) {
  print(stringr::str_c("Realization ", i))
  individual_discrete(init, p = 0.3, r = 5) %>%
    print()
}
```



**(b) Continuous time individual-based modeling**

*(i) Write a function that simulates a fixed number of time units of a continuous time individual-based SIR model. Your function should have the following user specified arguments: initial vector of individual labels, infection rate $\beta$, removal rate $\gamma$, and the number of time units $t$.*

```{r}
individual_continuous <- function(init_labels, beta, gamma, t) {
  step_labs <- init_labels
  res <- matrix(init_labels, nrow = 1) %>% 
    as.data.frame() %>%
    magrittr::set_colnames(stringr::str_c("person_", 1:length(init_labels)))
  times <- c(0)
  repeat {
    i_t <- sum(step_labs == "I")
    tmp_df <- data.frame(state = step_labs) %>%
      dplyr::mutate(
        exp_race = purrr::map_dbl(
          .x = state,
          .f = ~{
            if (.x == "S") rexp(1, rate = beta * i_t)
            else if (.x == "I") rexp(1, rate = gamma)
            else if (.x == "R") NA_real_ 
          }
        ),
        state = dplyr::case_when(
          exp_race == min(exp_race, na.rm = TRUE) & state == "S" ~ "I",
          exp_race == min(exp_race, na.rm = TRUE) & state == "I" ~ "R",
          TRUE ~ state
        )
      )
    step_labs <- tmp_df$state
    res <- rbind(res, step_labs)
    times <- append(times, min(tmp_df$exp_race, na.rm = TRUE))
    
    if (!(sum(times, na.rm = TRUE) <= t & (!is.na(sum(times))))) break
  }
  res %>%
    magrittr::set_rownames(
      stringr::str_c("step_", 0:(nrow(.)-1))
    ) %>%
    dplyr::mutate(time_of_step = cumsum(times)) %>%
    dplyr::select(time_of_step, dplyr::everything()) %>%
    dplyr::filter(time_of_step < t)
}
```


*(ii) Assuming that initially the population of interest has 9 susceptible individuals, 1 infectious individual, and 0 removed individuals, print 5 realizations of label vectors, using $\beta = 0.8$, $\gamma = 1.5$, and $t = 2$. Print times of events and new label vectors at the event times.*

Note that these print-outs are in the reverse orientation as the ones in the previous parts, so that we could include a time variable as a column.

```{r}
init <- c("S", "S", "S", "S", "S", "S", "I", "S", "S", "S")

for (i in 1:5) {
  print(stringr::str_c("Realization ", i))
  individual_continuous(init, beta = 0.8, gamma = 1.5, t = 2) %>%
    print()
}
```


**(c) Continuous time compartmental modeling**

*(i) Write a function that simulates a fixed number of time units of a continuous time compartmental SIR model. Your function should have the following user specified arguments: initial vector of S, I, and R counts, infection rate $\beta$, removal rate $\gamma$, and the number of time units $t$.*

```{r}
compartment_continuous <- function(init_counts, beta, gamma, t) {
  times <- c(0)
  res <- data.frame(
    time = 0,
    S = init_counts[1],
    I = init_counts[2],
    R = init_counts[3]
  )
  step <- 1
  while (sum(times) < t) {
    i <- rexp(1, rate = beta * res$S[step] * res$I[step])
    r <- rexp(1, rate = gamma * res$I[step])
    if (!is.na(i) & !is.na(r) & i < r) {
      new_res <- data.frame(
        time = i + sum(times),
        S = res$S[step] - 1,
        I = res$I[step] + 1,
        R = res$R[step]
      )
      times <- append(times, i)
    } else if (!is.na(r)) {
      new_res <- data.frame(
        time = r + sum(times),
        S = res$S[step],
        I = res$I[step] - 1,
        R = res$R[step] + 1
      )
      times <- append(times, r)
    } else {
      break
    }
    step <- step + 1
    res <- rbind(res, new_res)
  }
  res %>%
    rbind(
      data.frame(
        time = t,
        S = res$S[nrow(res)],
        I = res$I[nrow(res)],
        R = res$R[nrow(res)]
      )
    )
}
```


*(ii) Assuming that initially the population of interest has 990 susceptible individuals, 10 infectious individuals, and 0 removed individuals, plot 5 realizations of SIR trajectories using $\beta = 0.008$, $\gamma = 3.5$, and $t = 4$.*

```{r, fig.height=4}
library(ggplot2)
init_labels <- c(990, 10, 0)

for (i in 1:5) {
  p <- compartment_continuous(init_labels, beta = 0.008, gamma = 3.5, t = 4) %>%
    tidyr::pivot_longer(
      cols = c("S", "I", "R"),
      names_to = "label",
      values_to = "value"
    ) %>%
    dplyr::mutate(
      label = dplyr::case_when(
        label == "S" ~ "Susceptible",
        label == "I" ~ "Infectious",
        label == "R" ~ "Removed"
      )
    ) %>%
    ggplot() +
    aes(x = time, y = value, color = label) +
    geom_line() +
    scale_color_manual(values = c("red", "black", "blue")) +
    xlab("Time") +
    ylab("Number of People") +
    ggtitle(
      stringr::str_c("Realization ", i, " of Compartmental SIR"),
      "Beta = 0.008, Gamma = 3.5, t = 4"
    ) +
    labs(color = "") +
    theme_bw() +
    theme(legend.position = c(0.8, 0.5))
  
  print(p)
}
```


## Problem 2

Formulate an extension of an SIR ODE model with vaccinated and unvaccinated compartments. Vaccinated individuals should be able to get infected and to transmit the disease.

**(a) Write down the differential equations**

First, we will define several functions based on the vaccinated vs. unvaccinated compartments:

- $V_S(t)$ is the number of susceptibles who are vaccinated
- $U_S(t)$ is the number of susceptibles who are unvaccinated
- $V_I(t)$ is the number of infectious who are vaccinated
- $U_I(t)$ is the number of infectious who are unvaccinated
- $V_R(t)$ is the number of removed who are vaccinated
- $U_R(t)$ is the number of removed who are unvaccinated

This leads to the overall functions:

- $S(t) = V_S(t) + U_S(t)$ is the total number of susceptibles in the population
- $I(t) = V_I(t) + U_I(t)$ is the total number of infectious in the population
- $R(t) = V_R(t) + U_R(t)$ is the total number of removed in the population

We can write the following differential equations for the vaccinated/unvaccinated compartments:
$$
\begin{aligned}
\frac{dV_S(t)}{dt} &= -\beta_1V_S(t)U_I(t) - \beta_2V_S(t)V_I(t) \\
\frac{dU_S(t)}{dt} &= -\beta_3U_S(t)U_I(t) - \beta_4U_S(t)V_I(t) \\
\frac{dV_I(t)}{dt} &= \beta_1V_S(t)U_I(t) + \beta_2V_S(t)V_I(t) - \gamma_1V_I(t) \\
\frac{dU_I(t)}{dt} &= \beta_3U_S(t)U_I(t) + \beta_4U_S(t)V_I(t) - \gamma_2U_I(t) \\
\frac{dV_R(t)}{dt} &= \gamma_1V_I(t) \\
\frac{dU_R(t)}{dt} &= \gamma_2U_I(t)
\end{aligned}
$$

And for the overall counts, we see that since derivatives are additive, we have
$$
\begin{aligned}
\frac{dS(t)}{dt} &= \frac{dV_S(t)}{dt} + \frac{dU_S(t)}{dt} = -\beta_1V_S(t)U_I(t) - \beta_2V_S(t)V_I(t) -\beta_3U_S(t)U_I(t) - \beta_4U_S(t)V_I(t) \\
\frac{dI(t)}{dt} &= \frac{dV_I(t)}{dt} + \frac{dU_I(t)}{dt} \\
&= \beta_1V_S(t)U_I(t) + \beta_2V_S(t)V_I(t) - \gamma_1V_I(t) + \beta_3U_S(t)U_I(t) + \beta_4U_S(t)V_I(t) - \gamma_2U_I(t) \\
\frac{dR(t)}{dt} &= \frac{dV_R(t)}{dt} + \frac{dU_R(t)}{dt} = \gamma_1V_I(t) + \gamma_2U_I(t)
\end{aligned}
$$
**(b) Provide interpretations of all parameters**

We can interpret the parameters as follows:

- $\beta_1$ is the transmission rate from unvaccinated people to vaccinated people
- $\beta_2$ is the transmission rate from vaccinated people to vaccinated people
- $\beta_3$ is the transmission rate from unvaccinated people to vaccinated people
- $\beta_4$ is the transmission rate from vaccinated people to unvaccinated people
- $\gamma_1$ is the recovery rate of vaccinated individuals
- $\gamma_2$ is the recovery rate of unvaccinated individuals


**(c) What parameters would allow you to measure vaccine efficacy against infection and transmission if you were able to estimate them from the data?**

Estimating $\beta_1, \beta_2, \beta_3, \beta_4$ would allow you to measure vaccine efficacy against infection and transmission.
For instance, $\beta_1$ and $\beta_2$ give you information on how likely it is for vaccinated people to become infected, while $\beta_3$ and $\beta_4$ give you information on how likely it is for unvaccinated people to become infected.
Looking at it from the other way, $\beta_1$ and $\beta_3$ give you information on how likely it is that unvaccinated people transmit the disease, while $\beta_2$ and $\beta_4$ give you information on how likely it is that vaccinated people transmit the disease.
