---
title: "Homework 1"
author: "Spencer Matthews"
date: "10/11/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Problem 1

Complete the following programming exercises in any language of your choice.

**(a) Discrete time individual-based modeling**

*(i) Write a function that simulates a fixed number of generations of a discrete time individual-based SIR model. Your function should have the following user specified arguments: initial vector of individual labels, infection probability $p$, and number of generations $r$*

```{r}

```


*(ii) Assuming that initially the population of interest has 9 susceptible individuals, 1 infectious individual, and 0 removed individuals, print 5 realizations of label vectors at each generation using $p = 0.3$ and $r = 5$.*


**(b) Continuous time individual-based modeling**

*(i) Write a function that simulates a fixed number of time units of a continuous tim individual-based SIR model. Your function should have the following user specified arguments: initial vector of individual labels, infection rate $\beta$, removal rate $\gamma$, and the number of time units $t$.*

*(ii) Assuming that initially the population of interest has 9 susceptible individuals, 1 infectious individual, and 0 removed individuals, print 5 realizations of label vectors, using $\beta = 0.8$, $\gamma = 1.5$, and $t = 2$. Print times of events and new label vectors at the event times.*

**(c) Continuous time compartmental modeling**

*(i) Write a function that simulates a fixed number of time units of a continuous time compartmental SIR model. Your function should have the following user specified arguments: initial ector of S, I, and R counts, infection rate $\beta$, removal rate $\gamma$, and the number of time units $t$.*

*(ii) Assuming that initially the population of interest has 990 susceptible individuals, 10 infectious individuals, and 0 removed individuals, plot 5 realizations of SIR trajectories using $\beta = 0.008$, $\gamma = 3.5$, and $t = 4$.*

## Problem 2

Formulate an extension of an SIR ODE model with vaccinated and unvaccinated compartments. Vaccinated individuals should be able to get infected and to transmit the disease.

**(a) Write down the differential equations**

**(b) Provide interpretations of all parameters**

**(c) What parameters would allow you to measure vaccine efficacy against infection and transmission if you were able to estimate them from the data?**
